Methods to prioritise pop-up active transport infrastructure and their application in a national cycleway prioritisation tool

### 2020-06-22 article.Rmd
Final tweaks

<!-- This paper reports on methods developed to support the identification of pop-up active transport infrastructure. -->
In the context of reduced public transport capacity in the wake of the COVID-19 pandemic, governments are scrambling to enable walking and cycling while adhering to physical distancing guidelines.
A range of pop-up options exist, include road space reallocation, which represents a 'quick win' for cities with 'spare space' along continuous road sections that have high latent cycling potential.
We developed methods to condense the complexity of city networks down to the most promising roads for road space reallocation schemes.
The resulting Rapid Cycleway Prioritisation Tool has been deployed for all Local Authorities in England to help prioritise emergency funds for new cycleways nationwide.
The methods and concepts could be used to support investment in pop-up infrastructure in cities worldwide.

@@

```{r gsub, fig.height=3, fig.cap="Illustration of geographic subsetting based on administrative boundaries (left), distance to a central point (middle) and distance to city centre or key destinations (right). Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
r_main_region_key = r_main_region[city_key_buffer, ]
r_main_city = r_main_region[region, ]
r_main_city$maxspeed[r_main_city$maxspeed == "20 mph or less"] = "20 mph (32.2 kmph) or less"
r_main_city$maxspeed[r_main_city$maxspeed == "30 mph"] = "30 mph (48.3 kmph)"
r_main_city$maxspeed[r_main_city$maxspeed == "40+ mph"] = "40+ mph (64.4+ kmph)"
m0 = tm_shape(r_main_city) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_layout(legend.only = TRUE)
m1 = tm_shape(r_main_city) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(city_centre_buffer) + tm_borders() +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
m3 = tm_shape(r_main_region_key) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
# this take aaaages to render...
tmap_arrange(m0, m1, m2, m3, nrow = 1, widths = c(0.1, 0.3, 0.3, 0.3))
```

>>>>

<!-- This paper reports on methods developed to support the identification of pop-up active transport infrastructure. -->
In the context of reduced public transport capacity in the wake of the COVID-19 pandemic, governments are scrambling to enable walking and cycling while adhering to physical distancing guidelines.
A range of pop-up options exist, include road space reallocation, which represents a 'quick win' for cities with 'spare space' along continuous road sections that have high latent cycling potential.
We developed methods to condense the complexity of city networks down to the most promising roads for road space reallocation schemes.
The resulting Rapid Cycleway Prioritisation Tool has been deployed for all local and regional transport authorities in England to help prioritise emergency funds for new cycleways nationwide.
The methods and concepts could be used to support investment in pop-up infrastructure in cities worldwide.

@@

```{r gsub, fig.height=3, fig.cap="Illustration of geographic subsetting based on administrative boundaries (left), distance to a central point (middle) and distance to city centre or key destinations (right). Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
r_main_region_key = r_main_region[city_key_buffer, ]
r_main_city = r_main_region[region, ]
r_main_city$maxspeed[r_main_city$maxspeed == "20 mph or less"] = "20 mph"
# r_main_city$maxspeed[r_main_city$maxspeed == "30 mph"] = "30 mph (48.3 kmph)"
# r_main_city$maxspeed[r_main_city$maxspeed == "40+ mph"] = "40+ mph (64.4+ kmph)"
m0 = tm_shape(r_main_city) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_layout(legend.only = TRUE)
m1 = tm_shape(r_main_city) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(city_centre_buffer) + tm_borders() +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
m3 = tm_shape(r_main_region_key) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(region) + tm_borders(col = "black", lwd = 2) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
# this take aaaages to render...
tmap_arrange(m0, m1, m2, m3, nrow = 1, widths = c(0.1, 0.3, 0.3, 0.3))
```

### 2020-06-22 article.tex
Final tweaks

In the context of reduced public transport capacity in the wake of the COVID-19 pandemic, governments are scrambling to enable walking and cycling while adhering to physical distancing guidelines.
A range of pop-up options exist, include road space reallocation, which represents a `quick win' for cities with `spare space' along continuous road sections that have high latent cycling potential.
We developed methods to condense the complexity of city networks down to the most promising roads for road space reallocation schemes.
The resulting Rapid Cycleway Prioritisation Tool has been deployed for all cities England to help prioritise emergency funds for new cycleways nationwide.
The methods and concepts could be used to support investment in pop-up infrastructure in cities worldwide.

@@

Much attention has focused on the impacts of COVID-19 on long-distance travel patterns (e.g.~Iacus et al. 2020; Jittrapirom and Tanaksaranond 2020) but short distance travel patterns have also changed.
There has been a notable increase in cycling in some areas (Harrabin 2020) due to the need increased need for exercise close to home for mental and physical health (Jim\a'enez-Pav\a'on, Carbonell-Baeza, and Lavie 2020) and a reduction in public transport options (e.g.~Tian et al. 2020).
The second reason is particularly important given that many `key workers' are low paid, with limited access to private automobiles.

Local and national governments are working out how best to respond.
Many options are available to ensure that citizens can benefit from outdoor activity while minimising health risks, ranging from the hand sanitiser provision to the creation of extra active transport space (Freeman and Eykelbosh 2020).
Installation of `pop-up' active transport infrastructure has been endorsed and implemented in many places (Laker 2020).
The Scottish government, for example, has provided £10 million ``to keep key workers moving'' by ``reallocating road space to better enable this shift and make it safer for people who choose to walk, cycle or wheel for essential trips or for exercise'' (Transport Scotland 2020).
On 9\textsuperscript{th} May 2020, the UK government announced a £250 million package for pop-up active transport infrastructure (Reid 2020).
Significantly, alongside this funding comes updated \href{https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19}{statutory guidance} on pop-up infrastructure and safety.
Evidence is needed to ensure that such investment is spent effectively and where it is most needed.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Measures such as point closures or contraflow cycle lanes, which can be used to promote `filtered permeability', a strategy in which street networks are redesigned so that routes for cyclists are faster and more direct than routes for drivers. An example of this is \href{https://twitter.com/TowerHamletsNow/status/1257564043856019458}{shown} in {[}Tower Hamlets{]}.
\item
  Measures to close roads entirely to motor traffic or to particular classes of vehicle, either permanently or at certain times of day, as in New York's `Open Streets' scheme (Litman 2020).
\item
  Measures to reallocate space on wide roads to create new cycleways or to widen pavements (Orsman 2020).
\end{enumerate}

@@

The approach is not without limitations.
Its reliance on data rather than community engagement represents a rather top-down approach to transport planning.
To overcome this issue, future work could seek to incorporate the results such as those presented above into a participatory map of the type shown in Figure \ref{fig:commonplace}.
Further work could also extend the method in various ways, for example by refining estimates of cycling potential based on new parameters such as proximity to key destinations and estimates of road width.
We welcome feedback on the results and methods at \href{https://github.com/cyipt/popupCycleways}{github.com/cyipt/popupCycleways}.

>>>>

In the context of reduced public transport capacity in the wake of the COVID-19 pandemic, governments are scrambling to enable walking and cycling while adhering to physical distancing guidelines.
A range of pop-up options exist, include road space reallocation, which represents a `quick win' for cities with `spare space' along continuous road sections that have high latent cycling potential.
We developed methods to condense the complexity of city networks down to the most promising roads for road space reallocation schemes.
The resulting Rapid Cycleway Prioritisation Tool has been deployed for all local and regional transport authorities in England to help prioritise emergency funds for new cycleways nationwide.
The methods and concepts could be used to support investment in pop-up infrastructure in cities worldwide.

@@

Much attention has focused on the impacts of COVID-19 on long-distance travel patterns (e.g.~Iacus et al. 2020; Jittrapirom and Tanaksaranond 2020) but short distance travel patterns have also changed.
There has been a notable increase in cycling in some areas (Harrabin 2020) due to the increased need for exercise close to home for mental and physical health (Jim\a'enez-Pav\a'on, Carbonell-Baeza, and Lavie 2020) and a reduction in public transport options (e.g.~Tian et al. 2020).
The second reason is particularly important given that many `key workers' are low paid, with limited access to private automobiles.

Local and national governments are working out how best to respond.
Many options are available to ensure that citizens can benefit from outdoor activity while minimising health risks, ranging from hand sanitiser provision to the creation of extra active transport space (Freeman and Eykelbosh 2020).
Installation of `pop-up' active transport infrastructure has been endorsed and implemented in many places (Laker 2020).
The Scottish government, for example, has provided £10 million ``to keep key workers moving'' by ``reallocating road space to better enable this shift and make it safer for people who choose to walk, cycle or wheel for essential trips or for exercise'' (Transport Scotland 2020).
On 9\textsuperscript{th} May 2020, the UK government announced a £250 million package for pop-up active transport infrastructure (Reid 2020).
Significantly, alongside this funding comes updated \href{https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19}{statutory guidance} on pop-up infrastructure and safety.
Evidence is needed to ensure that such investment is spent effectively and where it is most needed.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Measures such as point closures or contraflow cycle lanes, which can be used to promote `filtered permeability', a strategy in which street networks are redesigned so that routes for cyclists are faster and more direct than routes for drivers. An example of this is \href{https://twitter.com/TowerHamletsNow/status/1257564043856019458}{shown} in {[}Tower Hamlets{]}.
\item
  Measures to close roads entirely to cars, either permanently or at certain times of day, as in New York's `Open Streets' scheme (Litman 2020).
\item
  Measures to reallocate space on wide roads to create new cycleways or to widen pavements (Orsman 2020).
\end{enumerate}

@@

The approach is not without limitations.
Its reliance on data rather than community engagement represents a rather top-down approach to transport planning.
The incorporation of our results into participatory maps such as the one presented in Figure \ref{fig:commonplace} will help to mitigate this limitation.
Further work could also extend the method in various ways, for example by refining estimates of cycling potential based on additional parameters such as proximity to key destinations.
We welcome feedback on the results and methods at \href{https://github.com/cyipt/popupCycleways}{github.com/cyipt/popupCycleways}.
